---
title: "Dependency ratios"
output: html_notebook
---

Trying to plot dependency ratios in terms of the traditional measure and also the kin measure. Issues: data availability. Based on the UN data, we can look at 15-64 year old women from 1990--2010 (even that is pushing it). This is five observations. 

Do all the code from last time to get some working data. 


```{r, message = F}
library(tidyverse)
library(here)
library(ggrepel)
source(here("code", "kin_functions.R"))
source(here("code", "alternative_daughters.R"))
source(here("code", "alternative_granddaughters.R"))
lt <- read_csv(here("data", "wpp_lt.csv"))
fx <- read_csv(here("data", "wpp_fx.csv"))
pop <- read_csv(here("data", "wpp_pop.csv"))


```

```{r}

df_mort <- lt %>% 
  # tidy up names
  rename(year = Period,
         region = `Region, subregion, country or area *`,
         age = `Age (x)`, 
         Lx = `Number of person-years lived L(x,n)`) %>% 
  rename(period = year) %>% 
  mutate(year = as.numeric(substring(period, 1,4))) %>% 
  # just keep the columns we need
  select(year, region, age, Lx)

# fertility df
df_fert <- fx %>% 
  # tidy up names
  rename(year = Period,
         region = `Region, subregion, country or area *`) %>% 
  # only keep years/ages/countries we want
  select(year, region, `15-19`:`45-49`) %>% 
  # change to long format
  gather(age_group, Fx, -year, -region) %>% 
  rename(period = year) %>% 
  mutate(year = as.numeric(substring(period, 1,4))) %>% 
    # make an age column 
  mutate(age = as.numeric(substring(age_group, 1, 2)))

df_pop <- pop %>% 
  # tidy up names
  rename(year = `Reference date (as of 1 July)`,
         region = `Region, subregion, country or area *`) %>% 
  select(year, region, `15-19`:`45-49`) %>% 
  # change to long format
  gather(age_group, pop, -year, -region) %>% 
  # make an age column 
  mutate(age = as.numeric(substring(age_group, 1, 2)))

df <- df_mort %>% 
  left_join(df_fert) %>% 
  left_join(df_pop) %>% 
  mutate(age_group = paste(age, age+4, sep = "-"),
         pop = ifelse(is.na(pop), 0, pop),
         Fx = ifelse(is.na(Fx), 0, Fx)) 

df <- df %>% 
  group_by(year, region) %>% 
  mutate(prop = pop/sum(pop))

```

```{r}
countries <- c("United States of America", 
               "Japan", "Nigeria", 
               "Venezuela (Bolivarian Republic of)",
               "Kenya",
               "Australia", 
               "Indonesia")
years <- c(1990, 1995, 2000, 2005, 2010)

df_data <- df %>% 
  filter(region %in% countries)

Ws <- df %>% ungroup() %>% filter(region %in% countries) %>% select(region, age, prop, year)
```

```{r}
df_res <- c()

for(i in 1:length(countries)){
  df_country <- c()
  for(j in 1:length(years)){
    this_df_data <- df_data %>% ungroup() %>% filter(region == countries[i])
    this_d_for_mothers <- df_data %>% ungroup() %>% filter(region == countries[i], year == years[j])
    this_W <- Ws %>% filter(region == countries[i])
    df_direct <- tibble(mother_age = seq(15, 60, by = 5),
                        country = countries[i],
                        year = years[j])
    df_direct <- df_direct %>% 
      rowwise() %>% 
      mutate(daughters = surviving_daughters_mean_age(df = this_df_data,
                                                                 year = years[j], 
                                                                 age_a = mother_age, 
                                                                 mean_age = 25),
             granddaughters = surviving_granddaughters_mean_age(df = this_df_data,
                                                                 year = years[j], 
                                                                 age_a = mother_age, 
                                                                 mean_age = 25),
             mothers = surviving_mothers_notstable(this_d_for_mothers, this_W, age_a = mother_age, years[j]),
             grandmothers = surviving_grandmothers_notstable(this_d_for_mothers, this_W, age_a = mother_age, years[j])
             
             )
    
    df_country <- rbind(df_country, df_direct)
  }
  df_res <- rbind(df_res, df_country)
}
```

Now we have a dataframe with some countries and the expected number of kin by year. Based on definition in abstract, calculate the dependency ratios

```{r}
df_res <- df_res %>% 
    #mutate(kdr = ifelse(mother_age<45,1,mothers+grandmothers))
  mutate(kdr = ifelse(mother_age<45,
                      (daughters + granddaughters)/(1+mothers+grandmothers),
                      (daughters + granddaughters +mothers+grandmothers)))

kdrs <- df_res %>% 
  left_join(df_pop) %>% 
    group_by(country, year) %>% 
    summarise(kdr = sum(kdr*pop/sum(pop)))
```
Plot


```{r}
df_res %>% 
  filter(year==2010) %>% 
  ggplot(aes(mother_age, kdr)) + geom_line()+ facet_grid(~country)
```

```{r}
kdrs %>% 
  filter(country %in% c("Japan", "United States of America", "Kenya", "Nigeria") ) %>% 
  ggplot(aes(year, kdr, color = country)) + geom_line(lwd = 1.2) + geom_point(size = 2)+
  theme_bw(base_size = 14) + ggtitle("Kin dependency ratios, 1990-2010")
ggsave("../fig/kdr_ts.png", width = 6, height = 4)
```

Now let's calculate OADR at the population level

```{r}
wb <- readRDS("../data/world_bank/all_indicators.rds")
wb <- wb %>% 
  filter(cntry_lab %in% c(countries, "Venezuela, RB", "United States")) %>% 
  mutate(country = case_when(
    cntry_lab == "Venezuela, RB" ~ "Venezuela (Bolivarian Republic of)",
    cntry_lab == "United States" ~ "United States of America", 
    TRUE ~ cntry_lab
  )) %>% 
  filter(year %in% years)
```

```{r}
wb %>% 
  filter(ind=="tadr", year>1985, cntry %in% c("JPN", "KEN", "VEN", "NGA", "USA")) %>% 
  left_join(kdrs) %>% 
  select(-ind, -ind_lab, -cntry_lab) %>% 
  rename(tadr = value) %>% 
  #mutate(tadr = tadr/100) %>% 
  #gather(ind, value, -cntry, -country, -year) %>% 
  #ggplot(aes(year, value, color = ind)) + geom_line() + facet_wrap(~cntry, scales = "free")
  ggplot(aes(kdr, tadr, color = cntry, group = cntry)) + 
  #facet_wrap(~cntry, scales = "free") + geom_point(size = 4) + 
  geom_point(size = 3)  + geom_line() + 
  scale_color_viridis_d()  + geom_text_repel(aes(label = year)) + 
  ggtitle("Total (population level) and kin (individual level) dependency ratios \n1990-2010")+
  ylab("Total dependency ratio (%)") + xlab("Kin dependency ratio (number per person)") + 
  theme_bw(base_size = 16)+ ylim(c(35, 110))
ggsave("../fig/kdr_tadr_scatter_all.png", width = 10, height = 6)
```

Facet 

```{r}
wb %>% 
  filter(ind=="tadr", year>1985, cntry %in% c("JPN", "KEN", "NGA", "USA")) %>% 
  left_join(kdrs) %>% 
  select(-ind, -ind_lab, -cntry_lab) %>% 
  rename(tadr = value) %>% 
  #mutate(tadr = tadr/100) %>% 
  #gather(ind, value, -cntry, -country, -year) %>% 
  #ggplot(aes(year, value, color = ind)) + geom_line() + facet_wrap(~cntry, scales = "free")
  ggplot(aes(kdr, tadr, color = year)) + 
  facet_wrap(~cntry, scales = "free") + geom_point(size = 4) + geom_line() + 
  #geom_point(size = 3)  + 
  scale_color_viridis_c()  + #geom_text_repel(aes(label = year)) + 
  ggtitle("Total (population level) and kin (individual level) dependency ratios \n1990-2010")+
  ylab("Total dependency ratio (%)") + xlab("Kin dependency ratio (number per person)") + 
  theme_bw(base_size = 16)
ggsave("../fig/kdr_tadr_scatter_facet.png", width = 10, height = 6)
```


Do the different types of kin fro 2010 for different countries 

```{r}
df_res %>% 
  select(-kdr) %>% 
  mutate(country_label = case_when(
    country=="Venezuela (Bolivarian Republic of)" ~ "Venezuela",
    country=="United States of America" ~ "USA",
    TRUE ~ country
  )) %>% 
  gather(type, number, -mother_age, -country, -year, -country_label) %>% 
  filter(mother_age<85, year ==2010, country_label %in% c("USA", "Japan", "Nigeria", "Kenya")) %>% 
  ggplot(aes(mother_age, number, color = type)) + geom_line(lwd = 1.2) +
  facet_grid(~country_label) + 
  ylab("expected number") + xlab("age") +
  ggtitle("Expected number of surviving kin for women at different ages, 2010") + 
  theme_bw(base_size = 16) 
ggsave("../fig/kin_country_facet_2010.png", width = 10, height = 5)
```

```{r}
df_res %>% 
  select(-kdr) %>% 
  mutate(country_label = case_when(
    country=="Venezuela (Bolivarian Republic of)" ~ "Venezuela",
    country=="United States of America" ~ "USA",
    TRUE ~ country
  )) %>% 
  gather(type, number, -mother_age, -country, -year, -country_label) %>% 
  filter(mother_age<85, year ==2010, country_label %in% c("Indonesia", "Japan", "Nigeria")) %>% 
  ggplot(aes(mother_age, number, color = type)) + geom_line(lwd = 1.2) +
  facet_grid(~country_label) + 
  ylab("no. per woman") + xlab("age") +
  ggtitle("Expected number of surviving kin \nfor women at different ages, 2010") + 
  theme_bw(base_size = 14) 
ggsave("../fig/kin_country_facet_2010_three.png", width = 8, height = 4)
```


TFR v e0 with kdr plotted on top (this does not work)

```{r}
wb %>% filter(ind %in% c("tfr", "e0"), country %in% countries) %>% 
  select(-ind_lab) %>% 
  spread(ind, value) %>% 
  left_join(kdrs) %>% 
  ggplot(aes(tfr, e0, color = country)) + geom_point() + geom_line(aes(kdr))
```




