---
title: "Some neat examples with non-stable relations"
output: html_notebook
---

The aim of this notebook is to plot some examples using the non-stable relations of daughters, grand-daughters, mothers and grand-mothers. 

Note: these relations are non-stable only in the sense of fertility / age-distribution. In both generational directions, there are $L_x$ terms that also should be altered, but I haven't done that yet. I think most of the change from stable to non-stable probably comes from the fertility aspects anyway.

Load in all the data on fertility, mortality and population age structures:

```{r, message = F}
library(tidyverse)
library(here)
source(here("code", "kin_functions.R"))
source(here("code", "alternative_daughters.R"))
source(here("code", "alternative_granddaughters.R"))
lt <- read_csv(here("data", "wpp_lt.csv"))
fx <- read_csv(here("data", "wpp_fx.csv"))
pop <- read_csv(here("data", "wpp_pop.csv"))
```
Clean up the data

```{r}

df_mort <- lt %>% 
  # tidy up names
  rename(year = Period,
         region = `Region, subregion, country or area *`,
         age = `Age (x)`, 
         Lx = `Number of person-years lived L(x,n)`) %>% 
  rename(period = year) %>% 
  mutate(year = as.numeric(substring(period, 1,4))) %>% 
  # just keep the columns we need
  select(year, region, age, Lx)

# fertility df
df_fert <- fx %>% 
  # tidy up names
  rename(year = Period,
         region = `Region, subregion, country or area *`) %>% 
  # only keep years/ages/countries we want
  select(year, region, `15-19`:`45-49`) %>% 
  # change to long format
  gather(age_group, Fx, -year, -region) %>% 
  rename(period = year) %>% 
  mutate(year = as.numeric(substring(period, 1,4))) %>% 
    # make an age column 
  mutate(age = as.numeric(substring(age_group, 1, 2)))

df_pop <- pop %>% 
  # tidy up names
  rename(year = `Reference date (as of 1 July)`,
         region = `Region, subregion, country or area *`) %>% 
  select(year, region, `15-19`:`45-49`) %>% 
  # change to long format
  gather(age_group, pop, -year, -region) %>% 
  # make an age column 
  mutate(age = as.numeric(substring(age_group, 1, 2)))

df <- df_mort %>% 
  left_join(df_fert) %>% 
  left_join(df_pop) %>% 
  mutate(age_group = paste(age, age+4, sep = "-"),
         pop = ifelse(is.na(pop), 0, pop),
         Fx = ifelse(is.na(Fx), 0, Fx)) 

df <- df %>% 
  group_by(year, region) %>% 
  mutate(prop = pop/sum(pop))
df
```

Pick some ages, years and regions.

```{r}
countries <- c("United States of America", 
               "Japan", "Nigeria", 
               "Venezuela (Bolivarian Republic of)",
               "Kenya",
               "Australia")
years <- 2010

df_data <- df %>% 
  filter(region %in% countries)

Ws <- df %>% ungroup() %>% filter(region %in% countries) %>% select(region, age, prop, year)
```

```{r}
df_res <- c()

for(i in 1:length(countries)){
  df_country <- c()
  for(j in 1:length(years)){
    this_df_data <- df_data %>% ungroup() %>% filter(region == countries[i])
    this_d_for_mothers <- df_data %>% ungroup() %>% filter(region == countries[i], year == years[j])
    this_W <- Ws %>% filter(region == countries[i])
    df_direct <- tibble(mother_age = seq(15, 85, by = 5),
                        country = countries[i],
                        year = years[j])
    df_direct <- df_direct %>% 
      rowwise() %>% 
      mutate(daughters = surviving_daughters_mean_age(df = this_df_data,
                                                                 year = years[j], 
                                                                 age_a = mother_age, 
                                                                 mean_age = 25),
             granddaughters = surviving_granddaughters_mean_age(df = this_df_data,
                                                                 year = years[j], 
                                                                 age_a = mother_age, 
                                                                 mean_age = 25),
             mothers = surviving_mothers_notstable(this_d_for_mothers, this_W, age_a = mother_age, years[j]),
             grandmothers = surviving_grandmothers_notstable(this_d_for_mothers, this_W, age_a = mother_age, years[j])
             
             )
    
    df_country <- rbind(df_country, df_direct)
  }
  df_res <- rbind(df_res, df_country)
}
```



```{r}
df_res %>% 
  mutate(country_label = case_when(
    country=="Venezuela (Bolivarian Republic of)" ~ "Venezuela",
    country=="United States of America" ~ "USA",
    TRUE ~ country
  )) %>% 
  gather(type, number, -mother_age, -country, -year, -country_label) %>% 
  filter(year==2010, mother_age > 15, mother_age<85) %>% 
  ggplot(aes(mother_age, number, color = type)) + geom_line(lwd = 0.8) +
  facet_wrap(~country_label) + 
  ylab("expected number") + xlab("age") +
  ggtitle("Expected number of surviving kin, 2010") + 
  theme_bw(base_size = 16) 
```

